---
title: "Air Quality Modeling"
output: html_document
date: "2026-03-03"
editor_options: 
  chunk_output_type: console
---


```{r echo=F}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      prompt = FALSE,
                      error = TRUE,
                      message = FALSE,
                      cache = FALSE,
                      cache.lazy = FALSE,
                      eval = TRUE,
                      eval.after = "fig.cap")

library(tidyverse)
library(rFIA)
library(ggplot2)
library(terra)
library(sf)
library(ggmap)
#library(rgdal)
library(lme4)
library(performance)
library(ggeffects)
library(plotrix)

select <- dplyr::select

# setting my preferred ggplot2 theme
theme_set(theme_bw())
theme_update(text = element_text(size=16, color = "black"),
             panel.grid.major = element_blank(),
             panel.grid.minor = element_blank(),
             strip.background = element_blank(),
             panel.border=element_rect(linewidth=1.5))

```

## Summary

Here I adapt Mark's air quality modeling code for my own use, with an eye toward troubleshooting emissions data and workflows. In particular, Mark's methods for chunking the CONUS rasters up to pyromes and processing separately may be useful. I'm working from the code document he shared "MK_air_quality_modeling_flamstat_v8.R"


### load spatial base data
```{r}
pyromes <- vect("../../../SHARED_DATA/pyromes/Data/Pyromes_CONUS_20200206.shp")

w.pyromes <- pyromes[1:45,]

states <- sf::st_read("../../../base_spatialdata/state_boundaries/state_boundaries.shp")

```

### load conditional wildfire emissions rasters

These rasters are the output of successive ToFuD, FlamStat, and FVS runs. It's my understanding that these represent all "2021" conditional emissions -- see workflow diagram in figures folder. There are serious issues here regarding conditional emissions, in particular, lots of situations where treatment emissions exceed those of the untreated scenario.


```{r}

ce_baseline <- rast("../data/flamstat/emissions_rasters/C_Emissions_FlamStat_NT.tif")
ce_Rx_fire <- rast("../data/flamstat/emissions_rasters/C_Emissions_FlamStat_RxFire.tif")
ce_Tx_fire <- rast("../data/flamstat/emissions_rasters/C_Emissions_FlamStat_Mech.tif")

```

### load FSIM burn probabilities

**download this**
```{r}

fsim_bp <- rast("../../../SHARED_DATA/FSIM/RDS-2020-0016-2__BP_CONUS/BP_CONUS/BP_CONUS.tif")

```


### loop to get conditional emissions and difference scenarios

```{r}

i <- 1

t_0 <- Sys.time()

for(i in 1:nrow(w.pyromes)){
  time_1 <- Sys.time()
  # filter to just pyrome in question
  pyrome <- pyromes[i,] %>% project(crs("EPSG:5070"))
  
  # print out pyrome output
  print(paste(pyrome$PYROME, pyrome$NAME))
  
  # filter conditional emissions to just the pyrome boundary
  py_ce_baseline <- ce_baseline %>% crop(pyrome, mask = T)
  py_ce_Rx_fire <- ce_Rx_fire %>% crop(pyrome, mask = T)
  py_ce_Tx_fire <- ce_Tx_fire %>% crop(pyrome, mask = T)
  
  gc()
  
  # write out rasters
  writeRaster(py_ce_baseline, 
              paste0("../data/flamstat/by_pyrome/emissions/conditional/baseline/baseline_conditional_emission_pyrome_",
                     pyrome$PYROME,".tif"), overwrite = T)
  writeRaster(py_ce_Rx_fire, 
              paste0("../data/flamstat/by_pyrome/emissions/conditional/Rx_fire/Rxfire_conditional_emission_pyrome_",
                     pyrome$PYROME,".tif"), overwrite = T)
  writeRaster(py_ce_Tx_fire, 
              paste0("../data/flamstat/by_pyrome/emissions/conditional/Tx_fire/Txfire_conditional_emission_pyrome_",
                     pyrome$PYROME,".tif"), overwrite = T)
  
  # filter burn probability to just the pyrome boundary
  #py_bp <- fsim_bp %>% crop(pyrome, mask = T)
  
  gc()
  
  ## get difference from baseline to treated ##
  py_ce_diff_Rx_fire <- py_ce_Rx_fire  - py_ce_baseline
  py_ce_diff_Tx_fire <- py_ce_Tx_fire - py_ce_baseline
  
  gc()
  
  # write out rasters
  writeRaster(py_ce_diff_Rx_fire, 
              paste0("../data/flamstat/by_pyrome/emissions/conditional/differences/Rxfire_conditional_diff_pyrome_", 
                     pyrome$PYROME, ".tif"), overwrite = T)
  writeRaster(py_ce_diff_Tx_fire, 
              paste0("../data/flamstat/by_pyrome/emissions/conditional/differences/Txfire_conditional_diff_pyrome_", 
                     pyrome$PYROME, ".tif"), overwrite = T)
  
  rm(py_ce_baseline)
  rm(py_ce_Rx_fire)
  rm(py_ce_Tx_fire)
  rm(py_ce_diff_Rx_fire)
  rm(py_ce_diff_Tx_fire)
  gc()
  
  time_2 <- Sys.time()
  
  print(time_2 - time_1)
}

print(paste0("Total time: ", Sys.time()- t_0))

```

### loop to get expected emissions and difference scenarios

```{r}

i <- 1
time_0 <- Sys.time()

for(i in 1:nrow(w.pyromes)){
  time_1 <- Sys.time()
  # filter to just pyrome in question
  pyrome <- pyromes[i,]
  
  # print out pyrome output
  print(paste(pyrome$PYROME, pyrome$NAME))
  
  # read conditional rasters
  py_ce_baseline <- rast(paste0("../data/flamstat/by_pyrome/emissions/conditional/baseline/baseline_conditional_emission_pyrome_",pyrome$PYROME,".tif"))
  py_ce_Rx_fire <- rast(paste0("../data/flamstat/by_pyrome/emissions/conditional/Rx_fire/Rxfire_conditional_emission_pyrome_",pyrome$PYROME,".tif"))
  py_ce_Tx_fire <- rast(paste0("../data/flamstat/by_pyrome/emissions/conditional/Tx_fire/Txfire_conditional_emission_pyrome_",pyrome$PYROME,".tif"))
  
  gc()
  
  # filter burn probability to just the pyrome boundary
  py_bp <- fsim_bp %>% crop(pyrome, mask = T)
  
  gc()
  
  ## multiply conditional emissions by burn probability to get annual expected emissions ##
  py_expected_baseline <- py_ce_baseline * py_bp
  py_expected_Rx_fire <- py_ce_Rx_fire * py_bp
  py_expected_Tx_fire <- py_ce_Tx_fire * py_bp
  
  rm(py_ce_baseline)
  rm(py_ce_Rx_fire)
  rm(py_ce_Tx_fire)
  rm(py_bp)
  gc()
  
  # write out rasters
  writeRaster(py_expected_baseline,
              paste0("../data/flamstat/by_pyrome/emissions/expected/baseline/baseline_expected_emission_pyrome",
                     pyrome$PYROME,".tif"), overwrite = T)
  writeRaster(py_expected_Rx_fire,
              paste0("../data/flamstat/by_pyrome/emissions/expected/Rx_fire/Rxfire_expected_emission_pyrome",
                     pyrome$PYROME,".tif"), overwrite = T)
  writeRaster(py_expected_Tx_fire,
              paste0("../data/flamstat/by_pyrome/emissions/expected/Tx_fire/Txfire_expected_emission_pyrome",
                     pyrome$PYROME,".tif"), overwrite = T)
  
  gc()
  
  ## get difference from baseline to treated ##
  py_expected_diff_Rx_fire <- py_expected_Rx_fire - py_expected_baseline
  py_expected_diff_Tx_fire <- py_expected_Tx_fire - py_expected_baseline
  
  
  gc()
  # write out rasters
  writeRaster(py_expected_diff_Rx_fire,
              paste0("../data/flamstat/by_pyrome/emissions/expected/differences/Rxfire_expected_diff_pyrome_",
                     pyrome$PYROME, ".tif"), overwrite = T)
  writeRaster(py_expected_diff_Tx_fire,
              paste0("../data/flamstat/by_pyrome/emissions/expected/differences/Txfire_expected_diff_pyrome_",
                     pyrome$PYROME, ".tif"), overwrite = T)
  
  # remove files
  rm(py_expected_baseline)
  rm(py_expected_Rx_fire)
  rm(py_expected_Tx_fire)
  
  rm(py_expected_diff_Rx_fire)
  rm(py_expected_diff_Tx_fire)
  
  gc()
  
  time_2 <- Sys.time()
  
  print(time_2 - time_1)
}

print(paste0("Total time: ",Sys.time()-time_0))

```

### merge resulting pyrome-level conditional and expected emissions rasters

```{r}
# Run as background job -- rastermerge_gb1.R
# Conditional emissions

ce_baseline_list <- list.files("../data/flamstat/by_pyrome/emissions/conditional/baseline/", full.names=T)
ce_baseline <- sprc(ce_baseline_list) %>% merge()
writeRaster(ce_baseline, "../data/flamstat/emissions_rasters/dp_merged/conditional_baseline_conus.tif")
gc()
rm(ce_baseline)

ce_Rx_fire_list <- list.files("../data/flamstat/by_pyrome/emissions/conditional/Rx_fire/", full.names=T)
ce_Rx_fire <- sprc(ce_Rx_fire_list) %>% merge()
writeRaster(ce_Rx_fire, "../data/flamstat/emissions_rasters/dp_merged/conditional_Rx_fire_conus.tif")
gc()
rm(ce_Rx_fire)

ce_Tx_fire_list <- list.files("../data/flamstat/by_pyrome/emissions/conditional/Tx_fire/", full.names=T)
ce_Tx_fire <- sprc(ce_Tx_fire_list) %>% merge()
writeRaster(ce_Tx_fire, "../data/flamstat/emissions_rasters/dp_merged/conditional_Tx_fire_conus.tif")
gc()
rm(ce_Tx_fire)

ce_Rx_diff_list <- list.files("../data/flamstat/by_pyrome/emissions/conditional/differences/Rx*", full.names=T)
ce_Rx_diff <- sprc(ce_Tx_fire_list) %>% merge()
writeRaster(ce_Tx_diff, "../data/flamstat/emissions_rasters/dp_merged/conditional_Rx_fire_diff_conus.tif")
gc()
rm(ce_Rx_diff)

ce_Tx_diff_list <- list.files("../data/flamstat/by_pyrome/emissions/conditional/differences/Tx*", full.names=T)
ce_Tx_diff <- sprc(ce_Tx_fire_list) %>% merge()
writeRaster(ce_Tx_diff, "../data/flamstat/emissions_rasters/dp_merged/conditional_Tx_fire_diff_conus.tif")
gc()
rm(ce_Tx_diff)


# Expected emissions

exp_baseline_list <- list.files("../data/flamstat/by_pyrome/emissions/expected/baseline/", full.names=T)
exp_baseline <- sprc(exp_baseline_list) %>% merge()
writeRaster(exp_baseline, "../data/flamstat/emissions_rasters/dp_merged/expected_baseline_conus.tif")
gc()
rm(exp_baseline)

exp_Rx_fire_list <- list.files("../data/flamstat/by_pyrome/emissions/expected/Rx_fire/", full.names=T)
exp_Rx_fire <- sprc(exp_Rx_fire_list) %>% merge()
writeRaster(expexp_Rx_fire, "../data/flamstat/emissions_rasters/dp_merged/expected_Rx_fire_conus.tif")
gc()
rm(exp_Rx_fire)

exp_Tx_fire_list <- list.files("../data/flamstat/by_pyrome/emissions/expected/Tx_fire/", full.names=T)
exp_Tx_fire <- sprc(exp_Tx_fire_list) %>% merge()
writeRaster(exp_Tx_fire, "../data/flamstat/emissions_rasters/dp_merged/expected_Tx_fire_conus.tif")
gc()
rm(exp_Tx_fire)

exp_Rx_diff_list <- list.files("../data/flamstat/by_pyrome/emissions/expected/differences/Rx*", full.names=T)
exp_Rx_diff <- sprc(exp_Tx_fire_list) %>% merge()
writeRaster(exp_Tx_diff, "../data/flamstat/emissions_rasters/dp_merged/expected_Rx_fire_diff_conus.tif")
gc()
rm(exp_Rx_diff)

exp_Tx_diff_list <- list.files("../data/flamstat/by_pyrome/emissions/expected/differences/Tx*", full.names=T)
exp_Tx_diff <- sprc(exp_Tx_fire_list) %>% merge()
writeRaster(exp_Tx_diff, "../data/flamstat/emissions_rasters/dp_merged/expected_Tx_fire_diff_conus.tif")
gc()
rm(exp_Tx_diff)

```


### Can I use this same approach to compare e.g., flame length rasters?

```{r}

rx.fl <- rast("data/flamstat/flamelength_rasters/Rx_fire_FVS_flamelength_masked.tif")/1000

wfnt20_rx20 <- rast("data/flamstat/RH_comparisons/ConWFNT20_RxFire20_comp/ConWFNT20_RxFire20_comp.tif")

rx.fl <- rx.fl %>% crop(ext(wfnt20_rx20))

terra::plot(rx.fl, wfnt20_rx20, maxcells = 1000,
            pch = 19, rgb(0,0,0,0.5), cex = 1.5)

rx2 <- rx.fl
rx.fl[wfnt20_rx20>0] <- NA

x <- values(rx.fl,na.rm=T)

y <- rx.fl %>% 
  spatSample(., size=1e5, na.rm=T)

y %>% 
  ggplot(.,
         aes(x = Rx_fire_FVS_flamelength_masked)) + 
  geom_density()

```

